---
title: Analysis of human CART sorted from cured NSG mice
author: "M. Andreatta <massimo.andreatta at unil.ch> and S. Carmona <santiago.carmona at unil.ch>"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'splenic.Il10_CART.html'))})
#output: html_notebook
---

```{r}
renv::restore()

library(remotes)
library(Seurat)
library(ggplot2)

library(viridis)

#remotes::install_github("carmonalab/ProjecTILs", ref="v3.0.0")
#remotes::install_github("carmonalab/UCell", ref="e54df8e")
library(ProjecTILs)
library(UCell)
```

Set up directories
```{r}
#NB: download the data and set 'ddir' to your data directory
ddir <- "~/Dropbox/CSI/Collaborations/LiTang/CART_human/"
ddir_10x <- sprintf("%s/sc-RNA-seq_filtered_cellranger", ddir)  

list.files(ddir_10x)

#make plots directory
dir.create(file.path("plots"), showWarnings = FALSE)
```

```{r}
set.seed(123)
mat_il10 <- ReadMtx(mtx=sprintf("%s/matrix-group-IL10-CD19-CAR-T.mtx.gz",ddir_10x),
                    cells=sprintf("%s/barcodes-group-IL10-CD19-CAR-T.tsv.gz",ddir_10x),
                    features=sprintf("%s/features-group-IL10-CD19-CAR-T.tsv.gz",ddir_10x))

obj_il10 <- CreateSeuratObject(mat_il10, project = "human_IL10",
                               min.cells=1, min.features = 50)

obj_il10 <- RenameCells(obj_il10, add.cell.id = "IL10")
obj_il10$Sample <- "IL10"

mat_ctrl <- ReadMtx(mtx=sprintf("%s/matrix-group-CD19-CAR-T.mtx.gz",ddir_10x),
                    cells=sprintf("%s/barcodes-group-CD19-CAR-T.tsv.gz",ddir_10x),
                    features=sprintf("%s/features-group-CD19-CAR-T.tsv.gz",ddir_10x))

obj_ctrl <- CreateSeuratObject(mat_ctrl, project = "human_IL10",
                               min.cells=1, min.features = 50)

obj_ctrl <- RenameCells(obj_ctrl, add.cell.id = "CTRL")
obj_ctrl$Sample <- "CTRL"

gex <- merge(obj_il10, obj_ctrl)
table(gex$Sample)
```


Ribosomal and mitochondrial content
```{r}
patterns <- c("^RP[LS]","^MT-")
gex <- AddMetaData(gex, metadata = PercentageFeatureSet(gex, pattern = patterns[1]), col.name = "percent.ribo")
gex <- AddMetaData(gex, metadata = PercentageFeatureSet(gex, pattern = patterns[2]), col.name = "percent.mito")
```


```{r fig.height=3}
Idents(gex) <- "Sample"
VlnPlot(gex, features = c("nFeature_RNA", "nCount_RNA","percent.ribo","percent.mito"), ncol = 2, pt.size=0)
```

Check QC stats
```{r}
quantile(gex$nFeature_RNA, probs=c(0.01,0.02,0.05,0.1,0.5,0.90,0.95,0.99))
quantile(gex$nCount_RNA, probs=c(0.01,0.02,0.05,0.1,0.5,0.90,0.95,0.99))
```

```{r}
gex <- subset(gex, subset= nFeature_RNA > 300 &
                nFeature_RNA < 5000 &
                nCount_RNA > 500 &
                nCount_RNA < 30000 &
                percent.mito < 20)

```


Load in list of genes to exclude from dim reduction (ribosomal, mitochondrial, TCR, cell cycle genes) that may distort the space
```{r}
if (!require("SignatuR", quietly = TRUE))
    install_github("carmonalab/SignatuR")

library(SignatuR)
print(SignatuR$Hs)

my.genes.blocklist <- c(GetSignature(SignatuR$Hs$Programs),
                        GetSignature(SignatuR$Hs$Compartments))
```


```{r fig.height=3}
set.seed(12345)
ndim=15

gex.list <- SplitObject(gex, split.by = "Sample")

gex.list <- lapply(gex.list, function(x) {
  x <- NormalizeData(x)
  
  x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 800, verbose = FALSE)
  VariableFeatures(x) <- setdiff(VariableFeatures(x), unlist(my.genes.blocklist))
  
  x <- ScaleData(x) |> RunPCA(features =VariableFeatures(x), npcs = ndim) |> RunUMAP(dims = 1:ndim, seed.use=123)
  x
})
```


```{r fig.height=3}
FeaturePlot(gex.list$IL10, features = c("SPI1","KLRB1","IGKC","APOE","CSF1R","C1QA"), ncol = 3)
FeaturePlot(gex.list$IL10, features = c("LCK","CD8A","CD8B","CD3E","CD4","MKI67"), ncol = 3)

FeaturePlot(gex.list$CTRL, features = c("SPI1","KLRB1","IGKC","APOE","CSF1R","C1QA"), ncol = 3)
FeaturePlot(gex.list$CTRL, features = c("LCK","CD8A","CD8B","CD3E","CD4","MKI67"), ncol = 3)
```

Evaluate multi-gene signatures
```{r fig.height=3, fig.width=5}
library(UCell)
library(patchwork)
signatures <- list("Tcell"=c("CD3D","CD3E","CD3G","CD2"),
                   "CD8T"=c("CD8A","CD8B"),
                   "CD4T"=c("CD4","CD40LG"),
                   "Bcell"=c("MS4A1","BANK1","PAX5","CD19"),
                   "MoMacDC"=c("LYZ2","CSF1R","MAFB","C1QA","APOE"),
                   "NK"=c("KLRD1","NKG7","NCR1","CD3D-","CD3E-"))

gex.list$IL10 <- AddModuleScore_UCell(gex.list$IL10, features = signatures, name = NULL)
gex.list$CTRL <- AddModuleScore_UCell(gex.list$CTRL, features = signatures, name = NULL)

p1 <- FeaturePlot(gex.list$IL10, features = names(signatures), 
                 max.cutoff = 'q99', pt.size = 0.5, combine=F)

wrap_plots(p1, ncol=3)

p2 <- FeaturePlot(gex.list$CTRL, features = names(signatures), 
                 max.cutoff = 'q99', pt.size = 0.5, combine=F)
wrap_plots(p2, ncol=3)
```


Use scGate to isolate CD8 T cells
```{r}
library(scGate)

models <- scGate::get_scGateDB()

model.Tcells <- models$human$generic$CD8T

gex.list <- lapply(gex.list, function(x){
   scGate(x, model=model.Tcells, reduction = "pca")
})

a <- DimPlot(gex.list$IL10, group.by = "is.pure") + ggtitle("IL10") + theme(aspect.ratio = 1)
b <- DimPlot(gex.list$CTRL, group.by = "is.pure") + ggtitle("CTRL") + theme(aspect.ratio = 1)
a | b

ggsave("plots/hCART_scGate_Tcell.png", height=3.5, width=7)
```
See why cells were removed
```{r fig.height=5}
feats <- grep("UCell",colnames(gex.list$IL10[[]]), value = T)

feats.show <- feats[1:12]

FeaturePlot(gex.list$IL10, features = feats.show, max.cutoff = 'q99')
ggsave("plots/hCART_CD8_broad_signatures.IL10.png", height=9, width=15)

FeaturePlot(gex.list$CTRL, features = feats.show, max.cutoff = 'q99')
ggsave("plots/hCART_CD8_broad_signatures.CTRL.png", height=9, width=15)
```

Quite a large fraction of gamma-delta T cells.
```{r}
hist(gex.list$CTRL$Tgammadelta_UCell, breaks=10)
hist(gex.list$IL10$Tgammadelta_UCell, breaks=10)
```


Filter T cells
```{r}
tcells <- lapply(gex.list, function(x){subset(x, subset=is.pure=="Pure")})
tcells
```

```{r fig.height=3}
set.seed(12345)
ndim=15

tcells <- lapply(tcells, function(x) {
  x <- NormalizeData(x)
  
  x <- FindVariableFeatures(x, selection.method = "vst", nfeatures = 800, verbose = FALSE)
  VariableFeatures(x) <- setdiff(VariableFeatures(x), unlist(my.genes.blocklist))
  
  x <- ScaleData(x) |> RunPCA(features =VariableFeatures(x), npcs = ndim) |> RunUMAP(dims = 1:ndim, seed.use=123)
  x
})
```


```{r fig.height=3, fig.width=6}
genes <- c("CD8A","CD4","SELL","IL7R","LEF1","CXCR3","ID3",
           "GZMK", "EOMES","TCF7","FOXP3","TOX","ID2","GZMA","GZMB","PDCD1","HAVCR2","LAG3","CXCL13","MKI67")

p1 <- FeaturePlot(tcells$IL10, features =genes, max.cutoff = 'q99',combine = FALSE)

for (i in seq_along(p1)) {
     p1[[i]] <- p1[[i]] + theme(aspect.ratio=1,
        axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        plot.title = element_text(size=12)) +
       scale_color_viridis(discrete = FALSE, option="C")
}
wrap_plots(p1, ncol=5)
ggsave("plots/hCART_CD8_umap_markers_pureTcells.IL10.png", height=9, width=17)

p2 <- FeaturePlot(tcells$CTRL, features =genes, max.cutoff = 'q99',combine = FALSE)

for (i in seq_along(p2)) {
     p2[[i]] <- p2[[i]] + theme(aspect.ratio=1,
        axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        plot.title = element_text(size=12)) +
       scale_color_viridis(discrete = FALSE, option="C")
}
wrap_plots(p2, ncol=5)
ggsave("plots/hCART_CD8_umap_markers_pureTcells.CTRL.png", height=9, width=17)
```

Evaluate multi-gene signatures
```{r fig.height=2.5, fig.width=8}
library(UCell)
library(patchwork)
signatures <- list("CD8T"=c("CD8A","CD8B"),
                   "CD4T"=c("CD4","CD40LG"),
                    "Cytotoxic"=c("PRF1","GZMA","GZMB"), 
                    "Stemness"= c("TCF7","CCR7","SELL","LEF1"),
                   "Exhaustion"=c("PDCD1","LAG3","HAVCR2","CTLA4"),
                   "Cycling"=c(my.genes.blocklist$cellCycle.G1S, my.genes.blocklist$cellCycle.G1S))

tcells$IL10 <- AddModuleScore_UCell(tcells$IL10, features = signatures, name = NULL)
tcells$CTRL <- AddModuleScore_UCell(tcells$CTRL, features = signatures, name = NULL)

p1 <- FeaturePlot(tcells$IL10, features = names(signatures), 
                 max.cutoff = 'q99', pt.size = 0.5, combine=F)


for (i in seq_along(p1)) {
     p1[[i]] <- p1[[i]] + theme(aspect.ratio=1,
        axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        plot.title = element_text(size=10)) +
       scale_color_viridis(discrete = FALSE, option="C")
}
wrap_plots(p1, ncol=6)

ggsave("plots/hCART_CD8_umap_signatures_pureTcells.IL10.png", height=3, width=16)


p2 <- FeaturePlot(tcells$CTRL, features = names(signatures), 
                 max.cutoff = 'q99', pt.size = 0.5, combine=F)
       
for (i in seq_along(p2)) {
     p2[[i]] <- p2[[i]] + theme(aspect.ratio=1,
        axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        plot.title = element_text(size=10)) +
       scale_color_viridis(discrete = FALSE, option="C")
}
wrap_plots(p2, ncol=6)

ggsave("plots/hCART_CD8_umap_signatures_pureTcells.CTRL.png", height=3, width=16)


```
Merge data
```{r}
tcells.merged <- merge(tcells$IL10, tcells$CTRL)
Idents(tcells.merged) <- "Sample"

VlnPlot(tcells.merged, features=names(signatures), pt.size = 0)

#Percent positive
100*sum(tcells$IL10$Stemness>0.2)/length(tcells$IL10$Stemness)
100*sum(tcells$CTRL$Stemness>0.2)/length(tcells$CTRL$Stemness)

```


Can we visualize the two conditions jointly
```{r}
set.seed(12345)
ndim=15

v1 <- VariableFeatures(tcells$IL10)
v2 <- VariableFeatures(tcells$CTRL)
vi <- intersect(v1, v2)

VariableFeatures(tcells.merged) <- vi

tcells.merged <- ScaleData(tcells.merged) |>
  RunPCA(features = vi, npcs = ndim) |>
  RunUMAP(dims = 1:ndim, seed.use=123)

palette.samples <- c("#ff7c79","#6596ff")
names(palette.samples) <- c("IL10","CTRL")

DimPlot(tcells.merged, group.by = "Sample", cols = palette.samples) + theme(aspect.ratio = 1)
ggsave("plots/hCART_CD8_umap_by_sample.pdf", height=4, width=5)
```

```{r fig.height=3, fig.width=6}
genes <- c("TCF7","IL7R","CCR7","SELL","LEF1","CXCR3","ID3",
           "GZMK","EOMES","FAS","TOX","ID2","GZMA","GZMB","PDCD1","HAVCR2","LAG3","MKI67")

p1 <- FeaturePlot(tcells.merged, features =genes, max.cutoff = 'q99',combine = FALSE)

for (i in seq_along(p1)) {
     p1[[i]] <- p1[[i]] + theme(aspect.ratio=1,
        axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        plot.title = element_text(size=12)) +
       scale_color_viridis(discrete = FALSE, option="C")
}
wrap_plots(p1, ncol=6)
ggsave("plots/hCART_CD8_markers_merged.png", height=7.5, width = 16)
ggsave("plots/hCART_CD8_markers_merged.pdf", height=7.5, width = 16)
```


```{r fig.height=3}
p1 <- FeaturePlot(tcells.merged, features = names(signatures), 
                 max.cutoff = 'q99', pt.size = 0.5, combine=F)

for (i in seq_along(p1)) {
     p1[[i]] <- p1[[i]] + theme(aspect.ratio=1,
        axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        plot.title = element_text(size=10)) +
       scale_color_viridis(discrete = FALSE, option="C")
}
wrap_plots(p1, ncol=3)
```

Smooth UCell scores
```{r fig.height=3}
tcells.merged <- SmoothKNN(tcells.merged, reduction="pca", k=10,
                           signature.names=names(signatures), suffix = "_kNN")

p1 <- FeaturePlot(tcells.merged, features = c("Stemness_kNN","Cytotoxic_kNN","Cycling_kNN"), 
                 pt.size = 0.5, combine=F)

for (i in seq_along(p1)) {
     p1[[i]] <- p1[[i]] + theme(aspect.ratio=1,
        axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        plot.title = element_text(size=10)) +
       scale_color_viridis(discrete = FALSE, option="C")
}
wrap_plots(p1, ncol=3)

ggsave("plots/hCART_CD8_signatures_kNN_merged.png", height=4, width=12)
ggsave("plots/hCART_CD8_signatures_kNN_merged.pdf", height=4, width=12)
```
```{R}
Idents(tcells.merged) <- "Sample"
VlnPlot(tcells.merged, features=c("Stemness_kNN","Cytotoxic_kNN","Cycling_kNN"),
        pt.size = 0, cols = palette.samples)

ggsave("plots/hCART_signatures_violins.pdf", height=5, width=8)

tcell.split <- SplitObject(tcells.merged)

#Percent positive
100*sum(tcell.split$IL10$Stemness_kNN>0.2)/ncol(tcell.split$IL10)
100*sum(tcell.split$CTRL$Stemness_kNN>0.2)/ncol(tcell.split$CTRL)

#Percent positive
100*sum(tcell.split$IL10$Cycling_kNN>0.1)/ncol(tcell.split$IL10)
100*sum(tcell.split$CTRL$Cycling_kNN>0.1)/ncol(tcell.split$CTRL)

```


Use ProjecTILs for annotation
```{r}
ref <- load.reference.map()

tcells$IL10 <- ProjecTILs.classifier(tcells$IL10, ref, filter.cells = F)
tcells$CTRL <- ProjecTILs.classifier(tcells$CTRL, ref, filter.cells = F)

palette <- c("#edbe2a", "#A58AFF", "#53B400", "#F8766D", 
                "#00B6EB", "#d1cfcc", "#FF0000", "#87f6a5", "#e812dd")

names(palette) <- levels(ref$functional.cluster)

DimPlot(tcells$IL10, reduction = "umap", group.by="functional.cluster", cols=palette) +
  ggtitle("Annotations from TIL reference") + theme(aspect.ratio = 1)

DimPlot(tcells$CTRL, reduction = "umap", group.by="functional.cluster", cols=palette) +
  ggtitle("Annotations from TIL reference") + theme(aspect.ratio = 1)

```
Composition barplot
```{r}
a <- plot.statepred.composition(ref=ref, query=tcells$IL10) + ggtitle("IL10")
b <- plot.statepred.composition(ref=ref, query=tcells$CTRL) + ggtitle("CTRL")

a | b
```

