---
title: Analysis of splenic CAR T cells
author: "M. Andreatta <massimo.andreatta at unil.ch> and S. Carmona <santiago.carmona at unil.ch>"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'splenic.Il10_CART.html'))})
#output: html_notebook
---

```{r}
renv::restore()

library(remotes)
library(Seurat)
library(ggplot2)

library(viridis)

#remotes::install_github("carmonalab/ProjecTILs", ref="v3.0.0")
library(ProjecTILs)
```

Set up directories
```{r}
#NB: set to your data directory
ddir <- "~/Dropbox/CSI/Collaborations/LiTang/CART_spleen/"
  
#make plots directory
dir.create(file.path("plots"), showWarnings = FALSE)
```

```{r}
data.seurat <- Read10X(sprintf("%s/data/10HCAR_matrix_10X_mouse", ddir))
dim(data.seurat)

gex <- CreateSeuratObject(data.seurat, project="Splenic_CART", min.cells = 3, min.features = 100)
gex 
```


Ribosomal and mitochondrial content
```{r}
patterns <- c("^Rp","^mt-")
gex <- AddMetaData(gex, metadata = PercentageFeatureSet(gex, pattern = patterns[1]), col.name = "percent.ribo")
gex <- AddMetaData(gex, metadata = PercentageFeatureSet(gex, pattern = patterns[2]), col.name = "percent.mito")
```


```{r fig.height=3}
Idents(gex) <- "Sample"
VlnPlot(gex, features = c("nFeature_RNA", "nCount_RNA","percent.ribo","percent.mito"), ncol = 2, pt.size=0)
```

Check QC stats
```{r}
quantile(gex$nFeature_RNA, probs=c(0.01,0.02,0.05,0.1,0.5,0.90,0.95,0.99))
quantile(gex$nCount_RNA, probs=c(0.01,0.02,0.05,0.1,0.5,0.90,0.95,0.99))
```

```{r}
gex <- subset(gex, subset= nFeature_RNA > 500 &
                nFeature_RNA < 4000 &
                nCount_RNA > 1000 &
                nCount_RNA < 10000 &
                percent.mito < 10)

```


Load in list of genes to exclude from dim reduction (ribosomal, mitochondrial, TCR, cell cycle genes) that may distort the space
```{r}
if (!require("SignatuR", quietly = TRUE))
    install_github("carmonalab/SignatuR")

library(SignatuR)
print(SignatuR$Mm)

my.genes.blocklist <- c(GetSignature(SignatuR$Mm$Programs),
                        GetSignature(SignatuR$Mm$Compartments))
```


```{r fig.height=3}
gex <- NormalizeData(gex)

gex <- FindVariableFeatures(gex, selection.method = "vst", nfeatures = 800, verbose = FALSE)
VariableFeatures(gex) <- setdiff(VariableFeatures(gex), unlist(my.genes.blocklist))
length(VariableFeatures(gex))
```

```{r}
set.seed(12345)
ndim=15

gex <- ScaleData(gex, do.scale = TRUE, do.center = TRUE) 
gex <- RunPCA(object = gex, features =VariableFeatures(gex), ndims.print = 1:5, nfeatures.print = 10, npcs = ndim)
gex <- RunUMAP(gex, reduction = "pca", dims = 1:ndim, seed.use=123)
```


```{r fig.height=3}
FeaturePlot(gex, features = c("Spi1","Tyrobp","Fcgr3","C1qb"))
FeaturePlot(gex, features = c("Vcan","Vcam1","Apoe","Csf1r"))
FeaturePlot(gex, features = c("Fn1","Mrc1","F13a1","C1qa"))
FeaturePlot(gex, features = c("Msr1","Mafb","Cd300e","Bank1"))
FeaturePlot(gex, features = c("S100a8","S100a9","Jchain","Lyz2"))
FeaturePlot(gex, features = c("Ms4a1","Cd74","Cd79b","Igkc"))
FeaturePlot(gex, features = c("Ptprc","Cd2","Cd3e","Cd3g"))
FeaturePlot(gex, features = c("Cd8a","Cd8b1","Sell","Tcf7"))
FeaturePlot(gex, features = c("Id3","Ccr7","Lef1","Il7r"))

FeaturePlot(gex, features = c("Spi1","Lyz2","S100a8","Ms4a1","Cd19","Cd8a"), ncol = 3)
ggsave("plots/splenic_CART_markers.pdf", height=6, width=12)
```

Evaluate multi-gene signatures
```{r fig.height=2, fig.width=6}
library(UCell)
library(patchwork)
signatures <- list("Tcell"=c("Cd3d","Cd3e","Cd3g","Cd2"), 
                   "Bcell"=c("Ms4a1","Bank1","Pax5","Cd19"),
                   "MoMacDC"=c("Lyz2","Csf1r","Mafb","C1qa","Apoe"),
                   "NK"=c("Klrd1","Nkg7","Ncr1","Cd3d-","Cd3e-","Cd3g-","Cd8a-","Cd8b1-"))

gex <- AddModuleScore_UCell(gex, features = signatures, name = NULL)

p <- FeaturePlot(gex, features = names(signatures), 
                 max.cutoff = 'q99', pt.size = 0.5, combine=F)

wrap_plots(p, ncol=4)

p <- p[1:3]       
for (i in seq_along(p)) {
     p[[i]] <- p[[i]] + theme(aspect.ratio=1.1,
        axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        plot.title = element_text(size=10)) +
       scale_color_viridis(discrete = FALSE, option="C")
}
wrap_plots(p, ncol=3)

ggsave("plots/splenic_CART_signatures_celltypes.pdf", height=4, width=10)


```

Use scGate to isolate T cells
```{r}
library(scGate)

models <- scGate::get_scGateDB()

model.Tcells <- models$mouse$generic$Tcell.alphabeta

#For some reason Ig chains are expressed. Remove Plasma cell filter
model.Tcells <- scGate::gating_model(model.Tcells, level=3, name="Plasma_cell", remove=T)
model.Tcells <- scGate::gating_model(model.Tcells, level=3, name="Macrophages", negative=T,
                                     signature = c("C1qa","C1qb","C1qc","Vcam1","Apoe"))

gex <- scGate(gex, model=model.Tcells, reduction = "pca")
DimPlot(gex) + ggtitle("scGate filter on T cells") + theme(aspect.ratio = 1)

ggsave("plots/splenic_CART_umap_Tcell_filter.pdf", height=3.5, width=7)
```
Filter T cells
```{r}
tcells <- subset(gex, subset=is.pure=="Pure")
```

```{r fig.height=3}
tcells <- NormalizeData(tcells)

tcells <- FindVariableFeatures(tcells, selection.method = "vst", nfeatures = 800, verbose = FALSE)
VariableFeatures(tcells) <- setdiff(VariableFeatures(tcells), unlist(my.genes.blocklist))


set.seed(12345)
ndim=15

tcells <- ScaleData(tcells, do.scale = TRUE, do.center = TRUE) 
tcells <- RunPCA(tcells, features =VariableFeatures(tcells), ndims.print = 1:5, nfeatures.print = 10, npcs = ndim)
tcells <- RunUMAP(tcells, reduction = "pca", dims = 1:ndim, seed.use=123)
```



```{r fig.height=3}
FeaturePlot(tcells, features = c("Igkc","Jchain","Ighm","Iglc3"), max.cutoff = 'q99')
FeaturePlot(tcells, features = c("Spi1","Tyrobp","Fcgr3","C1qb"), max.cutoff = 'q99')
FeaturePlot(tcells, features = c("Ikzf2","Fcer1g","Fcgr3","Tyrobp"), max.cutoff = 'q99')
FeaturePlot(tcells, features = c("Ms4a1","Cd74","Cd79b","Mki67"), max.cutoff = 'q99')
FeaturePlot(tcells, features = c("Ptprc","Cd2","Cd3e","Cd3g"), max.cutoff = 'q99')
FeaturePlot(tcells, features = c("Cd8a","Cd8b1","Cd4","Cd40lg"), max.cutoff = 'q99')
FeaturePlot(tcells, features = c("Tcf7","Ccr7","Lef1","Il7r"), max.cutoff = 'q99')
FeaturePlot(tcells, features = c("Tcell","Bcell","MoMacDC","NK"), max.cutoff = 'q99')
FeaturePlot(tcells, features = c("Prf1","Gzma","Gzmb","Gzmc"), max.cutoff = 'q99')
FeaturePlot(tcells, features = c("Ly6a","Cd69","Id2"), max.cutoff = 'q99')
```

```{r fig.height=3, fig.width=6}
p <- FeaturePlot(tcells, features = c("Cd8a","Cd4","Sell","Il7r","Lef1","Ccr7","Cxcr3","Id3",
                                "Ly6a","Eomes","Tcf7","Fas","Bcl2","Bcl6",
                                 "Tox","Id2","Gzma","Gzmb","Mki67","Top2a"),
                 max.cutoff = 'q99',combine = FALSE)

for (i in seq_along(p)) {
     p[[i]] <- p[[i]] + theme(aspect.ratio=1,
        axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        plot.title = element_text(size=12)) +
       scale_color_viridis(discrete = FALSE, option="C")
}
wrap_plots(p, ncol=5)

ggsave("plots/splenic_CART_umap_markers_pureTcells.pdf", height=9, width=17)
ggsave("plots/splenic_CART_umap_markers_pureTcells.png", height=9, width=17)
```
Show a smaller panel of genes
```{r}
p <- FeaturePlot(tcells, features = c("Sell","Lef1","Tcf7","Ccr7",
                                "Il7r","Ly6a","Id2","Id3",
                                 "Gzma","Gzmb","Mki67","Top2a"),
                 max.cutoff = 'q99',combine = FALSE)

for (i in seq_along(p)) {
     p[[i]] <- p[[i]] + theme(aspect.ratio=1,
        axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        plot.title = element_text(size=12)) +
       scale_color_viridis(discrete = FALSE, option="C")
}
wrap_plots(p, ncol=6)

ggsave("plots/splenic_CART_umap_markers_pureTcells.short.pdf", height=4, width=15)
ggsave("plots/splenic_CART_umap_markers_pureTcells.short.png", height=4, width=15)
```

Evaluate multi-gene signatures
```{r fig.height=2.5, fig.width=8}
library(UCell)
library(patchwork)
signatures <- list("Cytotoxic"=c("Prf1","Gzma","Gzmb","Gzmc","Gzmf"), 
                    "Stemness"= c("Tcf7","Ccr7","Sell","Lef1","Il7r"),
                   "Exhaustion"=c("Pdcd1","Lag3","Havcr2","Entpd1","Ctla4"),
                   "Memory"=c("Ly6a","Id2","Il7r"),
                   "Stemness_Memory"=c("Tcf7","Ccr7","Sell","Lef1","Il7r","Ly6a","Id2"),
                   "Cycling"=c(my.genes.blocklist$cellCycle.G1S, my.genes.blocklist$cellCycle.G1S))

tcells <- AddModuleScore_UCell(tcells, features = signatures, name = NULL)

p <- FeaturePlot(tcells, features = names(signatures), 
                 max.cutoff = 'q99', pt.size = 0.5, combine=F)

wrap_plots(p, ncol=5)
       
for (i in seq_along(p)) {
     p[[i]] <- p[[i]] + theme(aspect.ratio=1,
        axis.title=element_blank(),
        axis.text=element_blank(),
        axis.ticks=element_blank(),
        plot.title = element_text(size=10)) +
       scale_color_viridis(discrete = FALSE, option="C")
}
wrap_plots(p, ncol=3)

ggsave("plots/splenic_CART_umap_signatures_pureTcells.pdf", height=5, width=9)

```

Use ProjecTILs for annotation
```{r}
ref <- load.reference.map()

tcells <- ProjecTILs.classifier(tcells, ref, filter.cells = F)

palette <- c("#edbe2a", "#A58AFF", "#53B400", "#F8766D", 
                "#00B6EB", "#d1cfcc", "#FF0000", "#87f6a5", "#e812dd")

names(palette) <- levels(ref$functional.cluster)

DimPlot(tcells, reduction = "umap", group.by="functional.cluster", cols=palette) +
  ggtitle("Annotations from TIL reference") + theme(aspect.ratio = 1)

ggsave("plots/splenic_CART_umap_TILannotations.pdf", height=4, width=6)
```
Composition barplot
```{r}
plot.statepred.composition(ref=ref, query=tcells)
ggsave("plots/splenic_CART_barplot_TILannotations.pdf", height=3.5, width=5)
```


Using viral reference
```{r eval=F}
options(timeout = max(3000, getOption("timeout")))

refFileName <- "cd8_viral_reference.rds"
refUrl <- "https://figshare.com/ndownloader/files/23166794"
download.file(refUrl, refFileName)

ref.viral <- load.reference.map(refFileName)
ref.viral$viral.states <- ref.viral$functional.cluster
DefaultAssay(ref.viral) <- "integrated"

tcells <- ProjecTILs.classifier(tcells, ref.viral, filter.cells = F, labels.col = "viral.states")
DimPlot(tcells, reduction = "umap", group.by="viral.states") + ggtitle("Annotations from viral reference")

```



Unsupervised clustering
```{r}
kp <- 20
resol <- 0.2

set.seed(123)

tcells <- FindNeighbors(tcells, reduction = "pca", dims = 1:ndim, k.param = kp)

tcells  <- FindClusters(tcells, resolution = resol)
tcells$cluster <- tcells@active.ident

library(RColorBrewer)

display.brewer.pal(n = 8, name = 'Set2')
cpalette <- brewer.pal(n=length(unique(tcells$cluster))  , name="Set2")


g <- DimPlot(tcells, reduction = "umap", group.by = "cluster", pt.size = 0.5, cols=cpalette, label = T) + 
     NoLegend() + ggtitle("Unsupervised clusters") + theme(aspect.ratio=1)

g

```


Find DEG per cluster
```{r}
library(dplyr)
Idents(tcells) <- "cluster"

set.seed(1234)

cluster.markers <- FindAllMarkers(tcells, only.pos = T, min.pct = 0.1, min.diff.pct=0.1, 
                                  logfc.threshold = 0.25, max.cells.per.ident = 500, test.use="wilcox",base=exp(1))

all <- cluster.markers %>% group_by(cluster) %>% top_n(n = 50, wt = abs(avg_logFC))

for (i in levels(tcells@active.ident)) {
    print(subset(all, cluster==i))
}  

```
