---
title: Analysis of MC38-HER2 CAR-T cells, with or without IL10 treatment 
author: "M. Andreatta <massimo.andreatta at unil.ch> and S. Carmona <santiago.carmona at unil.ch>"
knit: (function(input_file, encoding) {
  out_dir <- 'docs';
  rmarkdown::render(input_file,
 encoding=encoding,
 output_file=file.path(dirname(input_file), out_dir, 'index.Il10_CART.html'))})
#output: html_notebook
---

```{r}
renv::activate()
renv::restore()

library(remotes)
library(Seurat)
library(ggplot2)
```


```{r}
data.seurat <- Read10X("~/Dropbox/CSI/Collaborations/LiTang/data/HER2_CAR_T")

#These are multiplexed data. Separate samples by their tag
names(data.seurat)
gex <- CreateSeuratObject(data.seurat[["Gene Expression"]], project="IL10_CART", min.cells = 3, min.features = 100)
gex 

valid.hashtags <- c("CMO309","CMO310")
valid.cells <- colnames(gex)
mpc <- data.seurat[["Multiplexing Capture"]][valid.hashtags,valid.cells]

gex[["hash"]] <- CreateAssayObject(counts = mpc)

gex <- NormalizeData(gex, assay = "hash", normalization.method = "CLR")

gex <- HTODemux(gex, assay = "hash", positive.quantile = 0.99)
```

Visualize demux results
```{r}
table(gex$hash.ID)
Idents(gex) <- "hash.ID"
RidgePlot(gex, assay = "hash", features = rownames(gex[["hash"]])[1:2], ncol = 2)
```
See demux assignments by CellRanger
```{r}
demux.CR <- read.csv("~/Dropbox/CSI/Collaborations/LiTang/multi/multiplexing_analysis/tag_calls_per_cell.csv", header=T)
rownames(demux.CR) <- demux.CR$cell_barcode

cells.use <- intersect(colnames(gex), demux.CR$cell_barcode)

gex@meta.data[cells.use,"sample.CR"] <- demux.CR[cells.use, "feature_call"]

table(gex$sample.CR, gex$hash.ID, useNA = "ifany")
gex
```


Visualize demux results
```{r}
#which annotation to use
gex$Sample <- gex$hash.ID
#gex$Sample <- gex$sample.CR

Idents(gex) <- "Sample"
RidgePlot(gex, assay = "hash", features = rownames(gex[["hash"]])[1:2], ncol = 2)

FeatureScatter(gex, feature1 = "hash_CMO309", feature2 = "hash_CMO310", group.by = "hash.ID") + ggtitle("By percentiles")
FeatureScatter(gex, feature1 = "hash_CMO309", feature2 = "hash_CMO310", group.by = "sample.CR") + ggtitle("By cellranger demux")
```



Ribosomal and mitochondrial content
```{r}
data.seurat <- gex
data.seurat <- AddMetaData(data.seurat, metadata = PercentageFeatureSet(data.seurat, pattern = "^Rp[ls]"), col.name = "percent.ribo")
data.seurat <- AddMetaData(data.seurat, metadata = PercentageFeatureSet(data.seurat, pattern = "^mt-"), col.name = "percent.mito")
```


```{r}
Idents(data.seurat) <- "Sample"
VlnPlot(data.seurat, features = c("nFeature_RNA", "nCount_RNA","percent.ribo","percent.mito"), ncol = 2, pt.size=0)
```

Filter out outliers & low quality cells
```{r}
cutoffs <- list()
cutoffs[["percent.ribo"]] <- c(min=max(quantile(data.seurat$percent.ribo,probs=c(0.01)),0),max=min(quantile(data.seurat$percent.ribo,probs=c(0.99)),60))

cutoffs[["percent.mito"]] <- c(min=max(quantile(data.seurat$percent.mito,probs=c(0.01)),0),max=min(quantile(data.seurat$percent.mito,probs=c(0.99)),10))

cutoffs[["nFeature_RNA"]] <- c(min=max(quantile(data.seurat$nFeature_RNA,probs=c(0.01)),500),max=min(quantile(data.seurat$nFeature_RNA,probs=c(0.99)),7000))
cutoffs[["nCount_RNA"]] <- c(min=max(quantile(data.seurat$nCount_RNA,probs=c(0.01)),200),max=min(quantile(data.seurat$nCount_RNA,probs=c(0.99)),50000))

print(cutoffs)

dim(data.seurat)

data.seurat <- subset(data.seurat, subset = nFeature_RNA > cutoffs[["nFeature_RNA"]]["min"] & nFeature_RNA < cutoffs[["nFeature_RNA"]]["max"] & 
                       nCount_RNA > cutoffs[["nCount_RNA"]]["min"] & nCount_RNA < cutoffs[["nCount_RNA"]]["max"] &
                       percent.ribo > cutoffs[["percent.ribo"]]["min"] &  percent.ribo < cutoffs[["percent.ribo"]]["max"] &
                    percent.mito > cutoffs[["percent.mito"]]["min"] & percent.mito < cutoffs[["percent.mito"]]["max"] )
dim(data.seurat)

Idents(data.seurat) <- "Sample"
VlnPlot(data.seurat, features = c("nFeature_RNA", "nCount_RNA","percent.ribo","percent.mito"), ncol = 2, pt.size=0)
```

```{r}
#exclude NA
cells <- colnames(data.seurat)[!is.na(data.seurat$Sample)]

if (length(cells)!=ncol(data.seurat)) {
   data.seurat <- subset(data.seurat, cells = cells)
}
#CMO309 corresponds to HER2_CART,  CMO309 corresponds to HER2_CART_IL10

data.seurat$condition <- as.character(data.seurat$Sample)
data.seurat$condition[data.seurat$Sample == "CMO309"] <- "WT"
data.seurat$condition[data.seurat$Sample == "CMO310"] <- "IL10"
Idents(data.seurat) <- "condition"
```

```{r}
library(scGate)
removeGenes <- unique(unique(unlist(scGate::genes.blacklist.default$Mm)))
length(removeGenes)
```


```{r}
data.seurat <- NormalizeData(data.seurat)
VlnPlot(data.seurat, features = c("Ptprc", "Cd2","Cd8a","Cd4"), ncol = 2, pt.size=0.01)
VlnPlot(data.seurat, features = c("Prf1", "Gzmb","Ifng","Pdcd1"), ncol = 2, pt.size=0.01)

data.seurat <- FindVariableFeatures(data.seurat, selection.method = "vst", nfeatures = 700, verbose = FALSE)
length(data.seurat@assays$RNA@var.features)
data.seurat@assays$RNA@var.features <- data.seurat@assays$RNA@var.features[!data.seurat@assays$RNA@var.features %in% removeGenes]
```


```{r}
set.seed(12345)
ndim=10

data.seurat <- ScaleData(data.seurat, do.scale = TRUE, do.center = TRUE) 
data.seurat <- RunPCA(object = data.seurat, features = data.seurat@assays$RNA@var.features, ndims.print = 1:5, nfeatures.print = 10, npcs = ndim)
data.seurat <- RunUMAP(data.seurat, reduction = "pca", dims = 1:ndim, seed.use=123)
```

```{r}
DimPlot(data.seurat, reduction = "umap", group.by = "condition") + ggtitle("UMAP by study")
```

```{r}
FeaturePlot(data.seurat, features = c("Spi1","Tyrobp","Fcgr3","Msr1"))
FeaturePlot(data.seurat, features = c("Ptprc","Cd2","Cd3e","Cd3g"))
FeaturePlot(data.seurat, features = c("Cd4","Cd8a","Cd8b1","Foxp3"))
```
Filter on T cells
```{r}
scGateModels <- get_scGateDB()

model.Tcell <- scGateModels$mouse$generic$Tcell
model.Tcell <- rbind(model.Tcell, c("level2","negative","Macrophage","Lyz2;C1qb;C1qc;Tyrobp"))

data.seurat <- scGate(data.seurat, model.Tcell, 
                      additional.signatures = list("G1S"=scGate::genes.blacklist.default$Mm$cellCycle.G1S,
                                                   "G2M"=scGate::genes.blacklist.default$Mm$cellCycle.G2M))
plot_levels(data.seurat)
data.seurat <- subset(data.seurat, subset=is.pure=="Pure")
```

Recalculate embeddings
```{r}
set.seed(12345)
ndim=10
data.seurat <- FindVariableFeatures(data.seurat, selection.method = "vst", nfeatures = 700, verbose = FALSE)
length(data.seurat@assays$RNA@var.features)
data.seurat@assays$RNA@var.features <- data.seurat@assays$RNA@var.features[!data.seurat@assays$RNA@var.features %in% removeGenes]

data.seurat <- ScaleData(data.seurat, do.scale = TRUE, do.center = TRUE) 
data.seurat <- RunPCA(object = data.seurat, features = data.seurat@assays$RNA@var.features, ndims.print = 1:5, nfeatures.print = 10, npcs = ndim)
data.seurat <- RunUMAP(data.seurat, reduction = "pca", dims = 1:ndim, seed.use=123)

DimPlot(data.seurat, reduction = "umap", group.by = "condition") + ggtitle("UMAP by condition")

data.seurat <- subset(data.seurat, subset=condition %in% c("WT","IL10"))

```

```{r fig.height=3}
palette2 <- c("whitesmoke","firebrick3")
FeaturePlot(data.seurat, features = c("Spi1","Tyrobp","Fcgr3","Msr1"), cols=palette2)
FeaturePlot(data.seurat, features = c("Ptprc","Cd2","Cd3e","Cd3g"), cols=palette2)
FeaturePlot(data.seurat, features = c("Cd4","Cd8a","Cd8b1","Foxp3"), cols=palette2)
FeaturePlot(data.seurat, features = c("Pdcd1","Havcr2","Prf1","Junb"), cols=palette2)

FeaturePlot(data.seurat, features = c("nCount_RNA","nFeature_RNA","percent.ribo","percent.mito"), cols=palette2)
```

Evaluate some signatures
```{r fig.height=2.5}
signatures <- list("Cytotoxic"=c("Prf1","Gzmb","Gzmc","Gzmk","Gzmg"), "Stemness"=c("Tcf7","Ccr7","Sell"),
                   "Exhaustion"=c("Pdcd1","Lag3","Havcr2","Entpd1"), 
                   "Cycling"=c(scGate::genes.blacklist.default$Mm$cellCycle.G1S, scGate::genes.blacklist.default$Mm$cellCycle.G2M))

data.seurat <- AddModuleScore_UCell(data.seurat, features = signatures)

FeaturePlot(data.seurat, features = c("Cytotoxic_UCell","Stemness_UCell","Exhaustion_UCell","Cycling_UCell"), cols=palette2, max.cutoff = 'q99', min.cutoff = 'q1')
```

Split plots
```{r fig.width=7, fig.height=2.5}
d1 <- subset(data.seurat, subset=condition == "WT")
p1 <- FeaturePlot(d1, features = c("Cytotoxic_UCell","Stemness_UCell","Exhaustion_UCell","Cycling_UCell"), cols=palette2, max.cutoff = 'q99', min.cutoff = 'q1', ncol = 4)
d2 <- subset(data.seurat, subset=condition == "IL10")
p2 <- FeaturePlot(d2, features = c("Cytotoxic_UCell","Stemness_UCell","Exhaustion_UCell","Cycling_UCell"), cols=palette2, max.cutoff = 'q99', min.cutoff = 'q1', ncol = 4)

p1 / p2

```
Unsupervised clustering
```{r}
kp <- 20
resol <- 0.4

set.seed(123)
data.seurat <- FindNeighbors(data.seurat, reduction = "pca", dims = 1:ndim, k.param = kp, nn.method = "rann")
  
data.seurat  <- FindClusters(data.seurat, resolution = resol)
data.seurat$cluster <- data.seurat@active.ident

library(scales)
nclusters <- length(levels(data.seurat$cluster))
cpalette <- hue_pal()(nclusters)

g <- DimPlot(data.seurat, reduction = "umap", group.by = "cluster", pt.size = 0.5, cols=cpalette, label = T) +
  ggtitle("Unsupervised clusters") + theme(aspect.ratio=1.2)
plot(g)

table(data.seurat$condition, data.seurat$cluster)
```

Find markers
```{r}
library(dplyr)
Idents(data.seurat) <- "cluster"

set.seed(1234)

cluster.markers <- FindAllMarkers(data.seurat, only.pos = T, min.pct = 0.1, min.diff.pct=0.1, 
                                  logfc.threshold = 0.25, max.cells.per.ident = 500, test.use="wilcox",base=exp(1))

all <- cluster.markers %>% group_by(cluster) %>% top_n(n = 50, wt = abs(avg_logFC))

for (i in levels(data.seurat@active.ident)) {
    print(subset(all, cluster==i))
}  

```


There are no clear markers for subtypes, except from a cluster of cycling cells and one small cluster of residual contaminants (myeloids?)







